---
title: "GOPrep"
author: "Cas Baptist"
date: "3/7/2022"
output: html_document
---

### Introduction

set org = the organism you wish to test

The heirarchical clustering step needs a larger stack. Run this before executing:

`
ulimit -s 65533
`

### Prerequisite libraries

```{r Libraries, echo=TRUE}
suppressPackageStartupMessages({
    library(R.utils)
    library(dplyr)
    library(tidyr)
    library(data.table)
    library(RColorBrewer)
    library(gplots)
    library(edgeR)
    library(reshape2)
    library(Polychrome)
    library(dendextend)
    library(ggplot2)
    library(DESeq2)
    library(dynamicTreeCut)
    library(biomaRt)
    library(qdapTools)
    library(hrbrthemes)
    library(tibble)

})
source("functions.R")
```

### 10. Gene Ontology and Gene Data 

A binary matrix was constructed with the Entrez Gene IDs (NCBI gene IDs) as rownames and GO
IDs as the column names (wide format).
If a GO ID is associated with a Gene ID, the cell will equal to 1, otherwise it will be zero. 

Gene ontologies downloaded from Ensembl BioMart with biomaRt package.

```{r, gene_ontol1}
# biomart
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
orgGOlong <- getBM(
  attributes=c("ensembl_gene_id","go_id"), 
  mart = ensembl)
orgGOlong <- orgGOlong[which(orgGOlong$go_id !="" ),]

## GO presence absence matrix
# gene ids in first column. each row represents a gene (ENSG ID)
# each column represents a GO term eg GO:000002
orgGOmatrixwide <- cbind(orgGOlong[1], mtabulate(as.data.frame(t(orgGOlong[-1]))))
#aggregate to reduce database
orgGOmatrixwide <- aggregate(. ~ ensembl_gene_id,orgGOmatrixwide,sum)
saveRDS(orgGOmatrixwide, "orgGOmatrixwide.rds")

# GeneIDs that are in the gene count data but not in the GO matrix
agg <- readRDS("agg.rds")
diffCountGO <- setdiff(rownames(agg), orgGOmatrixwide$ensembl_gene_id)
# GeneIDs that are in the GO matrix but not in the  gene count data
diffGOCount <- setdiff(orgGOmatrixwide$ensembl_gene_id, rownames(agg))
# Remove the GeneIDs that are in the GO matrix but not in the gene count data
IndexRemoved <- which(orgGOmatrixwide$ensembl_gene_id %in% diffGOCount)
GO_table <- orgGOmatrixwide
GO_table <- GO_table[-IndexRemoved,]

# for consistency and memory usage
dim(GO_table)
head(rowSums(GO_table[,2:ncol(GO_table)]),50)

GO_table <- aggregate(. ~ ensembl_gene_id,GO_table,sum)
dim(GO_table)
GO_table[1:10,1:6]

# Save the new GO table
saveRDS(GO_table, "GO_table.rds")

```

### 11. Blind GO Terms (90-10 split)

10 per cent of GeneIDs will have zero Gene Ontology Annotations. This will serve as the training data.
After the model has been optimised, these IDs will be unblinded as the testing data.

```{r, blind_terms, echo=TRUE}
# Fraction of the total number of genes to be blinded
test_size <- floor(0.10 * nrow(GO_table))
## set the seed to make your partition reproducible
set.seed(42)
blinded_ind <- sample(seq_len(nrow(GO_table)), size = test_size)
GO_train <- GO_table

# blinded aggregates from ontologies - training data
GO_train[blinded_ind, 2:ncol(GO_train)] = 0
GO_train <- aggregate(. ~ ensembl_gene_id,GO_train,sum)
saveRDS(GO_train, "GO_train.rds")

# blinded ontologies - testing data
GO_test <- GO_table[blinded_ind,]
saveRDS(GO_test, "GO_test.rds")

# scaled aggregate - testing data
normAggScale <- readRDS("normAggScale.rds")
normAggScale_test <- normAggScale[rownames(normAggScale) %in% GO_test$ensembl_gene_id,]
saveRDS(normAggScale_test, "normAggScale_test.rds")
```

### 12. Data removal

```{r, data_removal2, echo=FALSE}
remove(orgData)
remove(orgGOlong)
```

### Mem usage

```{r,memusage}
sort( sapply( ls() , function(x) { object.size( get( x ) ) }  )  )
```

### Session Information 

```{r Session Info, echo=FALSE}
sessionInfo()
```
