---
title: "HumanImptationBlinded"
author: "Cas Baptist"
date: "3/22/2022"
output: html_document
---

***

##### INTRO
***

</br>

### Prerequisite libraries

```{r, libraries}

suppressPackageStartupMessages({
    library(R.utils)
    library(dplyr)
    library(tidyr)
    library(data.table)
    library(RColorBrewer)
    library(gplots)
    library(Polychrome)
    library(caret)
    library(ggplot2)
    library(reshape2)
    library(tools)
    library(igraph)
    library("purrr")
    library(stringr)
})

source("functions.R")

```

</br>

### Load Pre-requisite RDS files 

```{r, load1}

normAggScale <- "normAggScale.rds"
GO_table <-"GO_table.rds"
GO_train <- "GO_train.rds"
cutClustvalues <- "cutClustvalues.rds"
cutClustvalues_dynamic <- "cutClustvalues_dynamic.rds"

#Download data [temporary - only uncomment if DataPrep not functioning due to memory issue]
# # if statement checks if object exists
# if (!file.exists("Data/cutClustvalues")){
#     download.file("https://ziemann-lab.net/public/ncRNA/cutClustvalues.rds", 
#               destfile = "Data/cutClustvalues")}
# cutClustvalues <- fread("Data/cutClustvalues")

# Normalised, aggregated, and scaled RNASeq counts
normAggScale <- readRDS(normAggScale)
# Binary df for GO annotations 
GO_table <- readRDS(GO_table)
# From GO_table, selected randomly as training data
GO_train <- readRDS(GO_train)
# List of genes grouped in different cluster sizes
cutClustvalues_dynamic <- readRDS(cutClustvalues_dynamic)
cutClustvalues <- readRDS(cutClustvalues)

```

### 1. Coarse Sweep for Spearman + complete linkage 

The following code will perform a K-fold validation for the imputation 
process which uses Spearman method to measure the distance matrix and 
complete as linkage method during the clustering process. 

```{r, coarsesweep1}

# Coarse sweep of the following parameters: cluster total(2407,211,117,70,36) and thresholds (0.1 to 0.9). The comments next to the cluster size in cl_list is the minumum cluster size used in the dynamic tree cut. These values are saved in the cutClustvalues_dynamic variable. 

#cluster totals taken from cutClustvalues_dynamic variable and taken from 5-figure-summary

#DYNAMIC SYSTEM to be created post-protyping stages - currently functions sub-par on local host

#extract out sizes from previous work
cl_list <- names(cutClustvalues_dynamic)
cl_list <- cl_list[seq(1,length(cl_list),5)]

thresh_list <- c(0.1, 0.3, 0.5, 0.6)

start <- Sys.time()

sample_kfold <- optimise_impute(cl_list=cl_list,
  thresh_list=thresh_list, 
  cuttree_values=cutClustvalues_dynamic,
  counts=normAggScale,
  GO_annot=GO_train)

end <- Sys.time()
totalTime <- end - start
totalTime

saveRDS(sample_kfold, "kfold_dynamic_all.rds")

```

### 2. Summary of performace measures for Pearson + complete linkage

```{r, summary1}

dat_list <- lapply(sample_kfold, function(x) x[names(x)!="Index_folds"])

# List the all the mean predicton scores for each parameter 
kfold_dynamic_sum <- summary_Csweep(dat_list)
saveRDS(kfold_dynamic_sum, "kfold_dynamic_sum.rds")

```

### 3. Visualizations

```{r, vis1}

# cut tree value vs Total Clusters – line diagram
cuttree_totals_df <- function(cutClustvalues){
  df <- as.data.frame(as.numeric(names(cutClustvalues)))
  m <- list()
  for (i in 1:length(cutClustvalues)){
    x <- cutClustvalues[[i]][["Cut_value"]]
    m[[i]] <- x
  }
  df$CuttreeVal <- as.numeric(m)
  colnames(df)[1] <- "ClTot"
  png("Data/cuttreeVSclTotal.png",
    width = 5*300,        # 5 x 300 pixels
    height = 5*300,
    res = 300,            # 300 pixels per inch
    pointsize = 8)        # smaller font size

  ggplot(df, aes(x=CuttreeVal, y=ClTot)) +
    geom_line() + ylab("Total Number of Clusters") +
    xlab("Cuttree Values") +
    ggtitle("Cuttree Value vs Total Number of Clusters") +
    theme(plot.title = element_text(hjust = 0.5))

  dev.off()
}

cuttree_totals_df(cutClustvalues)

# Total genes per cluster  – violin plot
c_violin <- function(cutClustvalues){

  clTotals <- do.call(rbind.data.frame, sapply(cutClustvalues, "[[", 3))
  clTotals$ClusterSize <- as.factor(as.numeric(str_extract_all(rownames(clTotals), "^[:digit:]*(?=.)")))
  clTotals$ClusterID <- as.numeric(str_extract_all(rownames(clTotals), "(?=.)\\d+$"))
  colnames(clTotals)[1] <- "TotalGenes"
  rownames(clTotals) <- c()

  # Filter according to cluster size
  cluster20_100 <- clTotals[clTotals$ClusterSize %in% c(20,50,100),]
  cluster200_500 <- clTotals[clTotals$ClusterSize %in% c(200, 300, 500),]
  cluster800_1500 <- clTotals[clTotals$ClusterSize %in% c(800, 1000, 1500),]

  pdf("Data/Violinplots.pdf")

  ggplot(cluster20_100, aes(x=ClusterSize, y=TotalGenes, fill=ClusterSize)) +
    geom_violin(trim=FALSE) + ylab("Total Number of Genes") +
    xlab("Cluster Size") +
    ggtitle("Distibution of genes per cluster size (20 to 100)") +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_boxplot(width=0.1, fill="white") + theme_minimal() +
    scale_fill_brewer(palette="Blues")

  ggplot(cluster200_500, aes(x=ClusterSize, y=TotalGenes, fill=ClusterSize)) +
    geom_violin(trim=FALSE) + ylab("Total Number of Genes") +
    xlab("Cluster Size") +
    ggtitle("Distibution of genes per cluster size (200 to 500)") +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_boxplot(width=0.1, fill="white") + theme_minimal() +
    scale_fill_brewer(palette="Dark2")

  ggplot(cluster800_1500, aes(x=ClusterSize, y=TotalGenes, fill=ClusterSize)) +
    geom_violin(trim=FALSE) + ylab("Total Number of Genes") +
    xlab("Cluster Size") +
    ggtitle("Distibution of genes per cluster size (800 to 1500)") +
    theme(plot.title = element_text(hjust = 0.5)) +
     geom_boxplot(width=0.1, fill="white") + theme_minimal() +
    scale_fill_brewer(palette="RdBu")

  dev.off()
}

c_violin(cutClustvalues)


```

### Session Information 

```{r Session Info, echo=FALSE}

sessionInfo()

```
