---
title: "HumanImptationBlinded"
author: "Cas Baptist"
date: "3/22/2022"
output: html_document
---

##### Imputation of training data as extracted from DataPrep stage. 

Training data makes up a randomised 10% of all data from preparation and aggregation stages, and is imputed to predict total cluster values and distances. Visualisations will differ depending on the 10% of data taken from DataPrep, but should be a statistically significant result when training from a large datatset. 


### Prerequisite libraries

```{r, libraries}

suppressPackageStartupMessages({
    library(R.utils)
    library(dplyr)
    library(tidyr)
    library(data.table)
    library(RColorBrewer)
    library(gplots)
    library(Polychrome)
    library(caret)
    library(ggplot2)
    library(reshape2)
    library(tools)
    library(igraph)
    library(purrr)
    library(stringr)
})

source("functions.R")

```

</br>

### Load Pre-requisite RDS files 

```{r, load1}

## Download data [temporary - only uncomment if DataPrep Chunk 7b not functioning due to memory issue]
## if downloading data, need to follow dynamic process from DataPrep (set below)

# if (!file.exists("Data/cutClustvalues")){
#     download.file("https://ziemann-lab.net/public/ncRNA/cutClustvalues.rds",
#               destfile = "Data/cutClustvalues")}
# cutClustvalues <- fread("Data/cutClustvalues")


## Load relevant objects

# Normalised, aggregated, and scaled RNASeq log2 counts - taken from DataPrep
normAggLog <- readRDS("normAggLog.rds")
# Binary dataframe for GO annotations - taken from GOPrep
GO_table <- readRDS("GO_table.rds")
# From GO_table, selected randomly as 10% training data - taken from GOPrep
GO_train <- readRDS("GO_train.rds")
# List of genes grouped in different cluster sizes - taken from DataPrep
cutClustvalues_dynamic <- readRDS("cutClustvalues_dynamic.rds")

```

### 1. Coarse Sweep for Spearman + complete linkage 

The following code will perform a K-fold validation for the imputation process, using the Spearman method to measure the distance matrix and complete linkage method during the clustering process. 

```{r, coarsesweep1}

# Coarse sweep of the following parameters: 
#     (1) cluster total (taken from cluster cuts and aggregates), and;
#     (2) thresholds (0.1 to 0.9, variable). 
# The length in cl_list is the minimum cluster size used in the dynamic tree cut. These values are saved in the   
#     cutClustvalues_dynamic variable. (see functions.R for details)
# Cluster totals are taken from cutClustvalues_dynamic variable and optimised

# extract out clusters lengths from cutClustvalues_dynamic 
cl_list <- names(cutClustvalues_dynamic)
cl_list <- cl_list[seq(1,length(cl_list),5)]
if (length(cl_list) == 0) { stop("extraction error, check 'cutClustvalues_dynamic.rds' is loaded") }

# THRESHOLDS TO BE DYNAMICALLY OPTIMISED AT A LATER DATE
# Optimal thresholds (dynamically found from ###)
thresh <- c(0.1, 0.3, 0.5, 0.6)

# Sample k-fold Validation
start <- Sys.time()

sample_kfold <- optimise_impute(cl_list, thresh, cutClustvalues_dynamic, normAggLog, GO_train)

end <- Sys.time()

#
totalTime <- end - start
totalTime

saveRDS(sample_kfold, "kfold_dynamic_all.rds")

```

### 2. Summary of performace measures for Pearson + complete linkage

```{r, summary1}

dat_list <- lapply(sample_kfold, function(x) x[names(x)!="Index_folds"])

# List the all the mean predicton scores for each parameter 
kfold_dynamic_sum <- summary_Csweep(dat_list)
saveRDS(kfold_dynamic_sum, "kfold_dynamic_sum.rds")

```

### 3. Visualizations

```{r, visualisations_blinded}

# cut tree value vs Total Clusters – line diagram
cuttree_totals_df <- function(cutClustvalues_dynamic){
  df <- as.data.frame(as.numeric(names(cutClustvalues_dynamic)))
  m <- list()
  for (i in 1:length(cutClustvalues_dynamic)){
    x <- cutClustvalues_dynamic[[i]][["Cut_value"]]
    m[[i]] <- x
  }
  df$CuttreeVal <- as.numeric(m)
  colnames(df)[1] <- "ClTot"
  png("Data/cuttreeVSclTotal.png",
    width = 5*300,        # 5 x 300 pixels
    height = 5*300,
    res = 300,            # 300 pixels per inch
    pointsize = 8)        # smaller font size

  ggplot(df, aes(x=CuttreeVal, y=ClTot)) +
    geom_line() + ylab("Total Number of Clusters") +
    xlab("Cuttree Values") +
    ggtitle("Cuttree Value vs Total Number of Clusters") +
    theme(plot.title = element_text(hjust = 0.5))

  dev.off()
}

cuttree_totals_df(cutClustvalues_dynamic)


# Total genes per cluster  – violin plot
c_violin <- function(cutClustvalues){

  clTotals <- do.call(rbind.data.frame, sapply(cutClustvalues_dynamic, "[[", 3))
  clTotals$ClusterSize <- as.factor(as.numeric(str_extract_all(rownames(clTotals), "^[:digit:]*(?=.)")))
  clTotals$ClusterID <- as.numeric(str_extract_all(rownames(clTotals), "(?=.)\\d+$"))
  colnames(clTotals)[1] <- "TotalGenes"
  rownames(clTotals) <- c()

  # Filter according to cluster size
  cluster20_100 <- clTotals[clTotals$ClusterSize %in% c(20,50,100),]
  cluster200_500 <- clTotals[clTotals$ClusterSize %in% c(200, 300, 500),]
  cluster800_1500 <- clTotals[clTotals$ClusterSize %in% c(800, 1000, 1500),]

  pdf("Data/Violinplots.pdf")

  ggplot(cluster20_100, aes(x=ClusterSize, y=TotalGenes, fill=ClusterSize)) +
    geom_violin(trim=FALSE) + ylab("Total Number of Genes") +
    xlab("Cluster Size") +
    ggtitle("Distibution of genes per cluster size (20 to 100)") +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_boxplot(width=0.1, fill="white") + theme_minimal() +
    scale_fill_brewer(palette="Blues")

  ggplot(cluster200_500, aes(x=ClusterSize, y=TotalGenes, fill=ClusterSize)) +
    geom_violin(trim=FALSE) + ylab("Total Number of Genes") +
    xlab("Cluster Size") +
    ggtitle("Distibution of genes per cluster size (200 to 500)") +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_boxplot(width=0.1, fill="white") + theme_minimal() +
    scale_fill_brewer(palette="Dark2")

  ggplot(cluster800_1500, aes(x=ClusterSize, y=TotalGenes, fill=ClusterSize)) +
    geom_violin(trim=FALSE) + ylab("Total Number of Genes") +
    xlab("Cluster Size") +
    ggtitle("Distibution of genes per cluster size (800 to 1500)") +
    theme(plot.title = element_text(hjust = 0.5)) +
     geom_boxplot(width=0.1, fill="white") + theme_minimal() +
    scale_fill_brewer(palette="RdBu")

  dev.off()
}

c_violin(cutClustvalues)


```

### Mem usage

```{r,memusage}
sort( sapply( ls() , function(x) { object.size( get( x ) ) }  )  )
```

### Session Information 

```{r Session Info, echo=FALSE}

sessionInfo()

```
