---
title: "DataPrep"
author: "Cas Baptist"
date: "3/7/2022"
output: html_document
---

### Introduction

set org = the organism you wish to test

The heirarchical clustering step needs a larger stack. Run this before executing:

`
ulimit -s 65533
`

### Prerequisite libraries

```{r Libraries, echo=TRUE}
suppressPackageStartupMessages({
    library(R.utils)
    library(dplyr)
    library(tidyr)
    library(data.table)
    library(RColorBrewer)
    library(gplots)
    library(edgeR)
    library(reshape2)
    library(Polychrome)
    library(dendextend)
    library(ggplot2)
    library(DESeq2)
    library(dynamicTreeCut)
    library(biomaRt)
    library(qdapTools)
    library(hrbrthemes)
    library(tibble)

})
source("functions.R")
```

### 10. Gene Ontology and Gene Data 

A binary matrix was constructed with the Entrez Gene IDs (NCBI gene IDs) as rownames and GO
IDs as the column names (wide format).
If a GO ID is associated with a Gene ID, the cell will equal to 1, otherwise it will be zero. 

Location for downloads: http://current.geneontology.org/products/pages/downloads.html

Will investigate using Ensembl BioMart.

```{r, gene_ontol1}
# biomart
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
orgGOlong <- getBM(
  attributes=c("ensembl_gene_id","go_id"), 
  mart = ensembl)
orgGOlong <- orgGOlong[which(orgGOlong$go_id !="" ),]
# GO presence absence matrix
# gene ids in first column
# each row represents a gene (ENSG ID)
# each column represents a GO term eg GO:000002
orgGOmatrixwide <- cbind(orgGOlong[1], mtabulate(as.data.frame(t(orgGOlong[-1]))))
# Save the dataframe to an RDS file
saveRDS(orgGOmatrixwide, "orgGOmatrixwide.rds")
# GeneIDs that are in the gene count data but not in the GO matrix
agg <- readRDS("agg.rds")
diffCountGO <- setdiff(rownames(agg), orgGOmatrixwide$ensembl_gene_id)
# GeneIDs that are in the GO matrix but not in the  gene count data
diffGOCount <- setdiff(orgGOmatrixwide$ensembl_gene_id, rownames(agg))
# Remove the GeneIDs that are in the GO matrix but not in the gene count data
IndexRemoved <- which(orgGOmatrixwide$ensembl_gene_id %in% diffGOCount)
GOtable <- orgGOmatrixwide
GOtable <- GOtable[-IndexRemoved,]

# the data structure is weird in that genes are shown as multiple rows
dim(GOtable)
head(rowSums(GOtable[,2:ncol(GOtable)]),50)

# Save the new GO table
saveRDS(GOtable, "GOtable.rds")
# Remove unneeded data
remove(orgGOlong)
```

### 11. Blind GO Terms (90-10 split)

10 per cent of GeneIDs will have zero Gene Ontology Annotations. This will serve as the training data.
After the model has been optimised, these IDs will be unblided as the testing data.

```{r, blind_terms, echo=TRUE}
#Fraction of the total number of genes to be blinded
test_size <- floor(0.10 * nrow(GOtable))
## set the seed to make your partition reproducible
set.seed(42)
blinded_ind <- sample(seq_len(nrow(GOtable)), size = test_size)
GOtrain <- GOtable
GOtrain[blinded_ind, 2:ncol(GOtrain)] = 0
saveRDS(GOtrain, "GOtrain.rds")
GO_test <- GOtable[blinded_ind,]
saveRDS(GO_test, "GO_test.rds")
normAggScale <- readRDS("normAggScale.rds")
normAggScale_test <- normAggScale[rownames(normAggScale) %in% GO_test$GeneID,]
saveRDS(normAggScale_test, "normAggScale_test.rds")
```

### 12. Data removal

```{r, data_removal2, echo=FALSE}
remove(orgData)
```

### Mem usage

```{r,memusage}
sort( sapply( ls() , function(x) { object.size( get( x ) ) }  )  )
```

### Session Information 

```{r Session Info, echo=FALSE}
sessionInfo()
```
