---
title: "GOPrep"
author: "Cas Baptist"
date: "3/7/2022"
output: html_document
---

### Introduction

set org = the organism you wish to test

The heirarchical clustering step needs a larger stack. Run this before executing:

`
ulimit -s 65533
`

### Prerequisite libraries

```{r Libraries, echo=TRUE}
suppressPackageStartupMessages({
    library(R.utils)
    library(dplyr)
    library(tidyr)
    library(data.table)
    library(RColorBrewer)
    library(gplots)
    library(edgeR)
    library(reshape2)
    library(Polychrome)
    library(dendextend)
    library(ggplot2)
    library(DESeq2)
    library(dynamicTreeCut)
    library(biomaRt)
    library(qdapTools)
    library(hrbrthemes)
    library(tibble)

})
source("functions.R")
```

### 1.//10. Load 

Gene ontologies downloaded from Ensembl BioMart with biomaRt package.

```{r, gene_ontol1}

# load gene ontologies from Ensembl. with BioMart package and form matrix
# If Ensembl. unresponsive due to remote connection error, mirror databases will be read.
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
orgGOlong <- getBM(
  attributes=c("ensembl_gene_id","go_id"), 
  mart = ensembl)
orgGOlong <- orgGOlong[which(orgGOlong$go_id !="" ),]

```

### 2.//11. GO presence absence matrix 

A binary matrix (GO_table in wide format) is constructed with the Entrez Gene IDs (NCBI gene IDs) as rownames and GO IDs as the column names
If a GO ID is associated with a Gene ID, the cell will equal to 1, else it will be zero. 

``` {r, gene_ontol2}

# Convert matrix to wide format
#     Gene IDs in first column. Each row represents a gene (ENSG ID)
#     each column represents a GO term eg GO:000002
orgGOmatrixwide <- cbind(orgGOlong[1], mtabulate(as.data.frame(t(orgGOlong[-1]))))

#aggregate to reduce database size
orgGOmatrixwide <- aggregate(. ~ ensembl_gene_id,orgGOmatrixwide,sum)
saveRDS(orgGOmatrixwide, "orgGOmatrixwide.rds")

#AGGREGATION OF ONTOLOGIES

# GeneIDs that are in the gene count data but not in the GO matrix
agg <- readRDS("agg.rds")
diffCountGO <- setdiff(rownames(agg), orgGOmatrixwide$ensembl_gene_id)
# GeneIDs that are in the GO matrix but not in the  gene count data
diffGOCount <- setdiff(orgGOmatrixwide$ensembl_gene_id, rownames(agg))

# Remove the GeneIDs that are in the GO matrix but not in the gene count data
IndexRemoved <- which(orgGOmatrixwide$ensembl_gene_id %in% diffGOCount)

# Associate matrix with unlinked Gene IDs, remove 
GO_table <- orgGOmatrixwide
GO_table <- GO_table[-IndexRemoved,]

# for consistency and memory usage
dim(GO_table)
head(rowSums(GO_table[,2:ncol(GO_table)]),50)

GO_table <- aggregate(. ~ ensembl_gene_id,GO_table,sum)
dim(GO_table)
GO_table[1:10,1:6]

# Save the new GO table
saveRDS(GO_table, "GO_table.rds")

```

### 11. Blind GO Terms (90-10 split)

10 per cent of GeneIDs will have zero Gene Ontology Annotations. This will serve as the training data and analysed in ImputationBlinded.
After the model has been optimised, these IDs will be unblinded as the testing data.

```{r, blind_terms, echo=TRUE}
# Fraction of the total number of genes to be blinded
train_size <- floor(0.10 * nrow(GO_table))

## Blind terms to the fraction
set.seed(42)
blinded_ind <- sample(seq_len(nrow(GO_table)), size = train_size)

# Associate blinded ontologies into a training matrix - training data for ImputationBlinded
GO_train <- GO_table
GO_train[blinded_ind, 2:ncol(GO_train)] = 0

#aggregate data into simplified matrix
GO_train <- aggregate(. ~ ensembl_gene_id,GO_train,sum)
saveRDS(GO_train, "GO_train.rds")


## UNBLINDING FOR TESTING DATA

# blinded ontologies - testing data for Imputation
GO_test <- GO_table[blinded_ind,]
saveRDS(GO_test, "GO_test.rds")

# Associate scaled aggregates found in GO_test matrix - testing data for Imputation
normAggScale <- readRDS("normAggScale.rds")
normAggScale_test <- normAggScale[rownames(normAggScale) %in% GO_test$ensembl_gene_id,]
saveRDS(normAggScale_test, "normAggScale_test.rds")
```

### 12. Data removal

```{r, data_removal2, echo=FALSE}
remove(orgData)
remove(orgGOlong)
```

### Mem usage

```{r,memusage}
sort( sapply( ls() , function(x) { object.size( get( x ) ) }  )  )
```

### Session Information 

```{r Session Info, echo=FALSE}
sessionInfo()
```
