---
title: "ImputationBlinded"
author: "Cas Baptist"
date: "3/22/2022"
output: html_document
---

***

##### INTRO
***

</br>

### Prerequisite libraries
```{r Libraries, echo=TRUE}
suppressPackageStartupMessages({
    library(R.utils)
    library(dplyr)
    library(tidyr)
    library(data.table)
    library(RColorBrewer)
    library(gplots)
    library(Polychrome)
    library(tidyverse)
    library(caret)
    library(ggplot2)
    library(reshape2)
    library(tools)
    library (igraph)
})

source("functions.R")

```
</br>

### Load Pre-requisite RDS files 
```{r echo=TRUE}
normAggScale <- "normAggScale.rds"
GO_table <-"GO_table.rds"
GO_train <- "GO_train.rds"
#file_cuttree_values <- "cuttree_values_complete.rds"
cutClustvalues_dynamic <- "cutClustvalues_dynamic.rds"

# Normalised, aggregated, and scaled RNASeq counts
normAggScale <- readRDS(normAggScale)
# Binary df for GO annotations 
GO_table <- readRDS(GO_table)
# From GO_table, selected randomly as training data
GO_train <- readRDS(GO_train)
# List of genes grouped in different cluster sizes
cutClustvalues_dynamic <- readRDS(cutClustvalues_dynamic)
```

### 1. Coarse Sweep for Spearman + complete linkage 
The following code will perform a K-fold validation for the imputation process which uses Spearman method to measure the distance matrix and complete as linkage method during the clustering process. 

```{r eval=FALSE, include=FALSE}
# Coarse sweep of the following parameters: cluster total(304,139,93,54,30) and thresholds (0.1 to 0.9). The comments next to the cluster size in cl_list is the minumum cluster size used in the dynamic tree cut. These values are saved in the cuttree_values variable. 

cl_list <- c(304, #10
             139, #20
             93, #30
             54, #50
             30) #102
#thresh_list <- c(0.1,0.5, 0.7, 0.9)
thresh_list <- c(0.2,0.3,0.4,0.6, 0.8)

start <- Sys.time()

sample_kfold <- optimise_impute(cl_list, thresh_list, cutClustvalues_dynamic, normAggScale, GO_train)

end <- Sys.time()
totalTime <- end -start
totalTime

saveRDS(sample_kfold, "kfold_dynamic_all.rds")

}

```

### 2. Summary of performace measures for Pearson + complete linkage

```{r}

dat_list <- lapply(sample_kfold, function(x) x[names(x)!="Index_folds"])

# List the all the mean predicton scores for each parameter 
kfold_dynamic_sum <- summary_Csweep(dat_list)
saveRDS(kfold_dynamic_sum, "kfold_dynamic_sum.rds")

```

### 3. Visualizations
NOT FUNCTIONING CURRENTLY 
```{r}

# # cut tree value vs Total Clusters – line diagram
# cuttree_totals_df <- function(cuttree_values){
#   df <- as.data.frame(as.numeric(names(cuttree_values)))
#   m <- list()
#   for (i in 1:length(cuttree_values)){
#     x <- cuttree_values[[i]][["Cut_value"]]
#     m[[i]] <- x
#   }
#   df$CuttreeVal <- as.numeric(m)
#   colnames(df)[1] <- "ClTot"
#   png("Data/cuttreeVSclTotal.png",
#     width = 5*300,        # 5 x 300 pixels
#     height = 5*300,
#     res = 300,            # 300 pixels per inch
#     pointsize = 8)        # smaller font size
# 
#   ggplot(df, aes(x=CuttreeVal, y=ClTot)) +
#     geom_line() + ylab("Total Number of Clusters") +
#     xlab("Cuttree Values") +
#     ggtitle("Cuttree Value vs Total Number of Clusters") +
#     theme(plot.title = element_text(hjust = 0.5))
# 
#   dev.off()
# }
# 
# cuttree_totals_df(cuttree_values)
# 
# # Total genes per cluster  – violin plot
# c_violin <- function(cuttree_values){
# 
#   clTotals <- do.call(rbind.data.frame, sapply(cuttree_values, "[[", 3))
#   clTotals$ClusterSize <- as.factor(as.numeric(str_extract_all(rownames(clTotals), "^[:digit:]*(?=.)")))
#   clTotals$ClusterID <- as.numeric(str_extract_all(rownames(clTotals), "(?=.)\\d+$"))
#   colnames(clTotals)[1] <- "TotalGenes"
#   rownames(clTotals) <- c()
# 
#   # Filter according to cluster size
#   cluster20_100 <- clTotals[clTotals$ClusterSize %in% c(20,50,100),]
#   cluster200_500 <- clTotals[clTotals$ClusterSize %in% c(200, 300, 500),]
#   cluster800_1500 <- clTotals[clTotals$ClusterSize %in% c(800, 1000, 1500),]
# 
#   pdf("Data/Violinplots.pdf")
# 
#   ggplot(cluster20_100, aes(x=ClusterSize, y=TotalGenes, fill=ClusterSize)) +
#     geom_violin(trim=FALSE) + ylab("Total Number of Genes") +
#     xlab("Cluster Size") +
#     ggtitle("Distibution of genes per cluster size (20 to 100)") +
#     theme(plot.title = element_text(hjust = 0.5)) +
#     geom_boxplot(width=0.1, fill="white") + theme_minimal() +
#     scale_fill_brewer(palette="Blues")
# 
#   ggplot(cluster200_500, aes(x=ClusterSize, y=TotalGenes, fill=ClusterSize)) +
#     geom_violin(trim=FALSE) + ylab("Total Number of Genes") +
#     xlab("Cluster Size") +
#     ggtitle("Distibution of genes per cluster size (200 to 500)") +
#     theme(plot.title = element_text(hjust = 0.5)) +
#     geom_boxplot(width=0.1, fill="white") + theme_minimal() +
#     scale_fill_brewer(palette="Dark2")
# 
#   ggplot(cluster800_1500, aes(x=ClusterSize, y=TotalGenes, fill=ClusterSize)) +
#     geom_violin(trim=FALSE) + ylab("Total Number of Genes") +
#     xlab("Cluster Size") +
#     ggtitle("Distibution of genes per cluster size (800 to 1500)") +
#     theme(plot.title = element_text(hjust = 0.5)) +
#      geom_boxplot(width=0.1, fill="white") + theme_minimal() +
#     scale_fill_brewer(palette="RdBu")
# 
#   dev.off()
# }
# 
# c_violin(cuttree_values)


```

### Session Information 

```{r Session Info, echo=FALSE}
sessionInfo()
```
