---
title: "Visualisations"
author: "Cas Baptist"
date: '2022-09-02'
output: html_document
---

Additional visualisations for exploration // documentation - do not run Data Removal chunk in DataPrep.Rmd

### Prerequisite Libraries

```{r Libraries, echo=TRUE}

suppressPackageStartupMessages({
  library(ggplot2)
  library(GGally)
  library(dendextend)
  library(data.table)
  library(RColorBrewer)
  library(gplots)
}}

```

### Load prerequisite files

```{r, load_files, echo=TRUE}
cutClustvalues_dynamic <- readRDS("cutClustvalues_dynamic")
normAggScale <- readRDS("normAggScale")
correlate_distClustVcoph <- readRDS("correlate_distClustVcoph")
hClust <- readRDS("hClust")


```


### Cluster Visualisations


```{r, Clusters}

# Dynamic cluster lengths
# TMM normalisation // Spearman correlation

if (!exists("P100")){
  P100 <- createPalette(100,  c("#ffdb58", "#DAA520", "#8B0000"))}
clusterCols <- P100[1:cutClustvalues_dynamic]
colfunc <- colorRampPalette(c("mustard","goldenrod","darkred"))

png("Data/TMM_sp.png", width = 5*300, height = 5*300, res = 300, pointsize = 8)
cutClusterSideBar <- clusterCols[cutClustvalues_dynamic[["Cluster Values"]]]
heatmap(correlate, main="Gene Correlation (TMM, Spearman)",  Rowv=as.dendrogram(hClust),
         dendrogram="both", scale="col", col = colfunc(25), trace="none",
         RowSideColors= cutClustusterSideBar, margins = c(5,5))
dev.off()


#Normalisaed PASS samples

png("Data/heatmaps_PASSnormalised.png", 
  width = 5*300, height = 5*300, res = 300, pointsize = 8)          

heatmap.2(normAgg, main="org PASS Samples",  Rowv=as.dendrogram(hClust),
          dendrogram="both", scale="column", col = colfunc(25), trace="none", 
          RowSideColors= myClusterSideBar, margins = c(5,5))
dev.off()


#Cluster counts

seeds_df_cl <- mutate(seeds_df, cluster = dend)
count(seeds_df_cl,cluster)


```

### Dendogram


```{r, dendograms_all}

# Correlation vs. Cophenetic Distance 

png("Data/dendro_average.png", width = 5*300, height = 5*300, res = 300, pointsize = 8)
dend <- as.dendrogram(hClust)
dendcols <- color_branches(dend, h = 3)
plot(dend, main = paste0("Complete Method: Corr = ", signif(correlate_distClustVcoph)))
print(dend)
dev.off()


```

### Histograms


```{r, histograms_all}

# Distributions of samples

# create a 5 x 5 inch png image
png("Data/hist_normAggScale.png", width = 5*300, height = 5*300, res = 300, pointsize = 8)
hist(normAggScale, breaks = 10)


# log2 counts per million

png("Data/hist_normAggLog.png", width = 5*300, height = 5*300, res = 300, pointsize = 8)
hist(normAggLog, breaks = 10)


#Spearman Correlation of RNASeq Counts

png("Data/hist_spcorr.png", width = 5*300, height = 5*300, res = 300, pointsize = 8)
hist(correlate, breaks = 20, xlab = "Gene Correlation Values", 
  main = "Spearman Correlation of RNASeq Counts")
dev.off()

```

### Imputation


```{r, imputation_graphs}

colfunc <- colorRampPalette(c("white", "red"))

# create heatmap for input matrix
heatmap.2(as.matrix(input_mat), main="org Cluster 1 GO Terms", scale="none",
    col = colfunc(25), trace="none", 
    margins = c(5,5))

# create heatmap for imputed df
heatmap.2(as.matrix(impClust1df), main="org Cluster 1 Imputed GO Terms", 
    scale="none", col = colfunc(25), trace="none", 
    margins = c(5,5))

# create heatmap for subtraction matrix
heatmap.2(as.matrix(Clust1neg), main="org Cluster 1 GO Terms vs Imputed GO Terms", 
    scale="none", col = colfunc(25), trace="none", 
    margins = c(5,5))

```

# Further Exploration

```{r, exploratoray_graphs}

# Create data
data <- data.frame( var1 = 1:100 + rnorm(100,sd=20), v2 = 1:100 + rnorm(100,sd=27), v3 = rep(1, 100) + rnorm(100, sd = 1))
data$v4 = data$var1 ** 2
data$v5 = -(data$var1 ** 2)

# Check correlations (as scatterplots), distribution and print correlation coefficient
ggpairs(data, title="correlogram with ggpairs()")
dev.off()